### 2011-07-30 docs/pubs/EIM-2011/abstract.tex
snapshot copy -- will finish up rest later

\begin{abstract}
  Effective discovery and integration of ecological data within data
  management systems requires rich semantic information that can
  describe and relate the types of information contained within
  disparate data sets. Within the Semtools project, we have developed
  approaches for expressing and representing semantic annotations of
  data sets for supplementing attribute and data-level metadata with
  terms drawn from domain-specific ontologies. Annotations provide a
  formal mechanism that can be used together with reasoning systems
  to enhance existing data discovery and integration approaches.  In
  this paper we describe extensions to the Ecological Metadata
  Language (EML) and associated tools for storing and using semantic
  annotations.  Specifically, we describe new user interface
  components implemented within the Morpho metadata editor for
  capturing user-supplied semantic annotations, extensions to the
  Metacat system for storing and accessing annotations and
  corresponding OWL-DL ontologies, and a new API within Metacat that
  uses annotation metadata to provide concept-based search and
  integration of data sets.
%   The Data Manager paper from the last EIM included semantic
%   foreshadowing in the 'future directions' section and this is a
%   logical continuation of that work. Previously, we saw that the DM
%   library enabled data integration tasks provided we knew exactly
%   which data attributes from exactly which datasets were "compatible"
%   with each other. Here "compatible" means not simply structurally
%   similar (e.g. decimal numbers) but also conceptually similar
%   (e.g. ocean water temperature in C). By including rich semantic
%   annotations we initially show simple integration capabilities where
%   the subject domain of our data corpus is relatively constrained
%   (i.e. SBC-LTER) and the observational models represented are
%   minimally variable (i.e. context excluded). Semantically-enahnced
%   query capabilities facilitate this 'smart integration' in multiple
%   ways.
%   First, the corpus is refined through compound concept-based
%   queries that exercise OBOE-compatible ontological subsumption
%   hierarchies to locate roughly equivalent attributes across
%   datasets. Additional query constraints are defined in relation to
%   the precise observational model formally expressed in OBOE. Finally,
%   the actual data values are interrogated so that only those that fall
%   in the desired range are returned. From this resultant subset of
%   disparate data objects we create a single synthetic data product
%   that represents a 'materialization' of the observations. Technical
%   discussion will include high-level descriptions of the
%   infrastructure and tools employed with focus on our development in
%   the rapidly evolving semantic frontier where scalability and best
%   practices are particularly germane and equally elusive.
\end{abstract}

>>>>

\noindent
\begin{abstract}
  Effective discovery and integration of ecological data within data
  management systems requires rich semantic information that can
  describe and relate the types of information contained within
  disparate data sets. Within the Semtools project, we have developed
  approaches for expressing and representing semantic annotations of
  data sets for supplementing attribute and data-level metadata with
  terms drawn from domain-specific ontologies. Annotations provide a
  formal mechanism that can be used together with reasoning systems to
  enhance existing data discovery and integration approaches.  We
  describe extensions to the Ecological Metadata Language (EML) and
  associated tools for storing and using semantic annotations.
  Specifically, we describe new user interface components implemented
  within the Morpho metadata editor for capturing user-supplied
  semantic annotations, extensions to the Metacat system for storing
  and accessing annotations and corresponding OWL-DL ontologies, and a
  new API within Metacat that uses annotation metadata to provide
  concept-based search and integration of data sets.
%   The Data Manager paper from the last EIM included semantic
%   foreshadowing in the 'future directions' section and this is a
%   logical continuation of that work. Previously, we saw that the DM
%   library enabled data integration tasks provided we knew exactly
%   which data attributes from exactly which datasets were "compatible"
%   with each other. Here "compatible" means not simply structurally
%   similar (e.g. decimal numbers) but also conceptually similar
%   (e.g. ocean water temperature in C). By including rich semantic
%   annotations we initially show simple integration capabilities where
%   the subject domain of our data corpus is relatively constrained
%   (i.e. SBC-LTER) and the observational models represented are
%   minimally variable (i.e. context excluded). Semantically-enahnced
%   query capabilities facilitate this 'smart integration' in multiple
%   ways.
%   First, the corpus is refined through compound concept-based
%   queries that exercise OBOE-compatible ontological subsumption
%   hierarchies to locate roughly equivalent attributes across
%   datasets. Additional query constraints are defined in relation to
%   the precise observational model formally expressed in OBOE. Finally,
%   the actual data values are interrogated so that only those that fall
%   in the desired range are returned. From this resultant subset of
%   disparate data objects we create a single synthetic data product
%   that represents a 'materialization' of the observations. Technical
%   discussion will include high-level descriptions of the
%   infrastructure and tools employed with focus on our development in
%   the rapidly evolving semantic frontier where scalability and best
%   practices are particularly germane and equally elusive.
\end{abstract}

\vspace{3pt}
\noindent {\small{\bf{\em Keywords}---ontologies; annotation; data
  discovery and integration}}

### 2011-07-30 docs/pubs/EIM-2011/application.tex
snapshot copy -- will finish up rest later

\mypara{Concept query.} The semantic query interface (see
\figref{fig:metacat-query}) is implemented as a Web application over
Metacat that primarily supports locating data sets by how well their
observational models match the given criteria. The interface provides
\emph{structured} as opposed to \emph{unstructured}, i.e.,
keyword-based queries. In particular, query criteria given by users
largely mirror the structure of a semantic annotation in that
combinations of Entity, Characteristic, and Protocol are specified and
optionally compounded when increased precision is sought.

@@

\mypara{Data-level query.} For even more precise recall, the OBOE
model can be (partially) materialized (see above) during the query
stage after which a data range filter can be applied. Different
techniques are available for merging the annotation with the data that
it describes, but for performance reasons a hybrid approach has been
adopted in which preliminary search results from a structured query
are collated and only that subset is materialized. Because our corpus
is described using EML in conjunction with the annotation syntax, the
Data Manager Library \cite{leinfelder10:_metad_driven_approac_to_loadin} is used to load the described
raw data (into a relational database) while the annotation informs the
correct use of the Data Manager query and filtering features. For any
measurements that match the concept query criteria, we verify that
those measurements (e.g., attributes) contain data values within the
range specified in the initial semantic data query and return the data
packages that contain them (see \figref{fig:metacat-query}).

\mypara{Data integration.} The materialization routine used for
semantic data queries laid the groundwork for enabling data
integration. In addition to inspecting the data for values within a
range and returning the data sets that contain a match, the data
integration feature of the Semtools Web application goes further by
constructing a synthetic data product (table) that represents the
complete results of the query in terms of both the attributes and the
values that are returned. Each original data set may have very
different syntactic structures (e.g. column number, naming, order) but
could share a subset of attributes that are semantically compatible as
defined in accompanying annotations. These compatible, in-common
attributes become the facets and metrics in a synthetic data set; their
values filtered accordingly. \figref{fig:integration} further
illustrates the data integration support provided in the current
implementation of the Web application. Consider two data packages,
denoted by A and B in \figref{fig:integration}.  Annotations (denoted
C and D) are used to determine semantically equivalent data attributes
contained in the data sets (denoted by E and F). The attributes {\tt
  plot} and {\tt site} are considered equivalent measurements of the
characteristic Location; attributes {\tt weight} and {\tt wt} both map
to the same characteristic Mass. The Semantic Mediation API parses
each annotation and then computes an equivalence mapping among
attributes based on their corresponding annotations. The Data Manager
Library is then used to load the data sets and then query each data
set to produce and merge the final, synthetic result data set.

>>>>

\subsection{Concept Query}

The semantic query interface (see \figref{fig:metacat-query}) is
implemented as a Web application over Metacat that primarily supports
locating data sets by how well their observational models match the
given criteria. The interface provides \emph{structured} as opposed to
\emph{unstructured}, i.e., keyword-based queries. In particular, query
criteria given by users largely mirror the structure of a semantic
annotation in that combinations of Entity, Characteristic, and
Protocol are specified and optionally compounded when increased
precision is sought.

@@

\subsection{Data-Level Query}

For even more precise recall, the OBOE model can be (partially)
materialized (see above) during the query stage after which a data
range filter can be applied. Different techniques are available for
merging the annotation with the data that it describes, but for
performance reasons a hybrid approach has been adopted in which
preliminary search results from a structured query are collated and
only that subset is materialized. Because our corpus is described
using EML in conjunction with the annotation syntax, the Data Manager
Library \cite{leinfelder10:_metad_driven_approac_to_loadin} is used to
load the described raw data (into a relational database) while the
annotation informs the correct use of the Data Manager query and
filtering features. For any measurements that match the concept query
criteria, we verify that those measurements (e.g., attributes) contain
data values within the range specified in the initial semantic data
query and return the data packages that contain them (see
\figref{fig:metacat-query}).

\subsection{Data Integration}

The materialization routine used for semantic data queries laid the
groundwork for enabling data integration. In addition to inspecting
the data for values within a range and returning the data sets that
contain a match, the data integration feature of the Semtools Web
application goes further by constructing a synthetic data product
(table) that represents the complete results of the query in terms of
both the attributes and the values that are returned. Each original
data set may have very different syntactic structures (e.g. column
number, naming, order) but could share a subset of attributes that are
semantically compatible as defined in accompanying annotations. These
compatible, in-common attributes become the facets and metrics in a
synthetic data set; their values filtered
accordingly. \figref{fig:integration} further illustrates the data
integration support provided in the current implementation of the Web
application. Consider two data packages, denoted by A and B in
\figref{fig:integration}.  Annotations (denoted C and D) are used to
determine semantically equivalent data attributes contained in the
data sets (denoted by E and F). The attributes {\tt plot} and {\tt
  site} are considered equivalent measurements of the characteristic
Location; attributes {\tt weight} and {\tt wt} both map to the same
characteristic Mass. The Semantic Mediation API parses each annotation
and then computes an equivalence mapping among attributes based on
their corresponding annotations. The Data Manager Library is then used
to load the data sets and then query each data set to produce and
merge the final, synthetic result data set.

### 2011-07-30 docs/pubs/EIM-2011/framework.tex
snapshot copy -- will finish up rest later

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.49\textwidth]{images/oboe}
  \caption{The main classes and properties of the extensible
    observation ontology (OBOE). While shown using Unified Modeling Language (UML), the model is
    defined using OWL-DL.}
  \label{fig:oboe}
\end{figure}

\mypara{The OBOE Observational Model.}  \figref{fig:oboe} shows the
main modeling constructs of OBOE (see:
\url{http://ecoinformatics.org/oboe/oboe.1.0/oboe-core.owl}). An {\em
  observation} is made of an {\em entity} (e.g., biological organisms,
geographic locations, environmental features) and serves to group a
set of measurements together to form a single ``\emph{observation
  event}''. A \emph{measurement} assigns a value to a {\em
  characteristic} of the observed entity (e.g., the weight of a
plant), and can also include \emph{standards} (e.g., units as well as
standards for coded values) and collection \emph{protocols}. An
observation can occur within the surrounding \emph{context} of other
observations (e.g., as part of a temporal or spatial context), and
context may include a named relationship (e.g., ``partOf'',
``within'') that existed during the observation event. A key feature
of OBOE is that it allows properties (characteristics and
relationships) of entities to be asserted without being interpreted as
\emph{inherently} (i.e., {always}) true of the entity.  Depending on
the context in which the entity was observed or how the measurements
were performed, an entity's properties may take on different values.
OBOE allows RDF-style assertions about entities to be contextualized,
and thus different values can be assigned for the same entity under
distinct contexts, which is a crucial feature for modeling ecological
as well as many other types of scientific data
\cite{bowers08,mungall07:_repres_phenot_in_owl}. In addition, OBOE is
currently implemented as an OWL-DL ontology that can be easily used
with (or extended by) other ontologies for specifying domain-specific
types of entities, characteristics, measurement standards, protocols,
and relationships. For instance, the Semtools project has defined
specific OBOE extensions in collaboration with the Santa Barbara
Coastal Long-Term Ecological Research Project as well as through
ongoing collaborations with other projects, and general extensions
exist for OBOE that define a number of common entities, measurement
units, and corresponding physical characteristics.

@@

\mypara{Semantic Annotations.} A semantic annotation consists of two
parts: (i) a ``configuration'' of the observation model containing the
specific entities, characteristics, observations, measurements, and so
on (drawn from one or more domain ontologies) that appropriately
capture the semantics of the data set; and (ii) a mapping between the
attributes in the data set to specific measurements defined in the
model configuration. \figref{fig:kelp-mass-model} shows a high-level
example of an annotation defined for a simple Kelp sampling data
set. Here, the data set consists of five attributes (bottom of
\figref{fig:kelp-mass-model}). Each attribute is mapped to a specific
measurement type (where only the characteristic of each measurement
type is shown), and measurement types are organized into observations
of specific Kelp entities (shown of type ``Macrocystis''), temporal
points (denoted by date-times), and spatial locations (given as site
names). Each measurement associated with a Kelp observation is assumed
to have occurred within the site and during the given time as
specified by the context relationships.

@@

Semantic annotations can be used to facilitate discovery and
integration of heterogeneous data sets. For instance, combining
semantic annotations with OBOE, it is possible to discover data sets
based on searches expressed over types of observations and
measurements of interest. As simple examples, users can pose queries
such as ``\emph{find all data sets containing observations of Kelp
  entities}'' and ``\emph{find all data sets containing Mass
  measurements of Kelp}''.\footnote{Note that queries also can include
  contextual constraints, e.g., Mass measurements of Kelp within a
  specific spatial or temporal context.} Both of these queries would
return the example data set in \figref{fig:kelp-mass-model} since the
attribute {\tt WET} is linked to a WetMass measurement for observations
of Macrocystis (where WetMass is defined as a special kind, or
\emph{subclass} of Mass, and Macrocystis is defined as a subclass of
Kelp). Using semantic annotations in this way can help to increase
both query recall and precision over standard keyword-based approaches
\cite{berkley09:_improv_data_discov_for_metad}. In particular, by
defining terms as subclasses of other terms (e.g., Macrocystis as a
subclass of Kelp), term expansion can be used to increase the number
(recall) of data sets returned (where subclasses of query terms are
also searched). The precision of the result can be improved since
queries may specify the desired connections between terms (e.g.,
measurements of Mass for Kelp observations) as opposed to returning
all data sets that simply mention the terms but without any explicit
connections (i.e., where Mass was measured, but not for Kelp samples).
Annotations also help facilitate integration by allowing tools to
align data set attributes based on their declared measurement and
observation types.

@@

\mypara{The Semantic Mediation API.} The Semantic Mediation API
includes basic ontology management features, annotation manipulation
capabilities, and simple concept navigation and visualization
components. The API is intended to be a centralized toolkit for use in
multiple application contexts (on either client or server
deployments).  The Semantic Mediation API uses both the OWL API
\cite{owlapi} for ontology management services including ontology 
parsing, serialization, and simple class and property exploration as
well as the Pellet description-logic reasoner \cite{pellet} for
classification and exposing inferred axioms in source ontologies. The
inference services exposed through the Semantic Mediation API are used
in both discovery and integration features described below.
% is particularly powerful when utilizing OBOE Measurement 'templates'
% that express exactly what (Entity and Characteristic) can be
% observed and how (Standard and Protocol).
In our current Morpho and Metacat extensions, semantic annotations are
managed and stored automatically in an underlying, local relational
database. While it is also possible to use in-memory approaches for
storing and querying annotations, we found the overhead to be
prohibitive when large numbers of data sets are managed.

@@

\mypara{The Morpho Editor Plugin.}  The semantic-annotation editor
plugin for Morpho provides a front-end to the Semantic Mediation API
and allows data owners and curators to define annotations for existing EML data 
descriptions. The editor provides a simple ``fill-in-the-blank'' style
form-based interface with a searchable hierarchical concept selection
widget (see \figref{fig:morpho-annotation}). The plugin seamlessly
integrates with a standard Morpho installation and provides semantic
query capabilities for locating data packages, marking up data sets within a 
package using semantic annotations, and saving annotations locally or
to a shared repository where they can be discovered and explored by
other users. The annotation editor in Morpho allows a user to view the
data set being annotated as they fill in (by selecting an appropriate
ontology term) the characteristic, measurement standard, protocol, and
associated entity for each data set attribute. Users can also specify
whether an observation spans multiple columns, and can provide context
relationships between attributes (i.e., observations). The editor
provides a number of additional features including the ability to
view the entire annotation (similar to \figref{fig:kelp-mass-model})
and to specify additional mapping constraints for observations and
measurements.

\mypara{Metacat Query Extensions.}  The semantic plugin for Metacat
augments Metacat's existing metadata storage and search by allowing
annotations to be saved and queried alongside the metadata and data
that they annotate. In addition to traditional keyword and spatial
search criteria, the Metacat plugin allows semantic criteria to be
included where they may either increase query recall using
term-expansion (i.e., traversing the class subsumption hierarchy) or
refine the result set by limiting matches to data sets that contain
the specified observational model (e.g., combinations of
OBOE-compatible entity, characteristic, measurement standard, or
protocol concepts). The observational model can be leveraged further
by materializing the annotation and data artifact (via the Data
Manager Library \cite{leinfelder10:_metad_driven_approac_to_loadin})
into a fully instantiated OBOE model and inspecting (and querying
over) the observational values themselves.

>>>>

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.494\textwidth]{images/oboe}
  \caption{Main classes and properties of the extensible observation
    ontology (OBOE). While shown using the Unified Modeling Language (UML),
    the model is defined as an OWL-DL ontology.}
  \label{fig:oboe}
  \vspace{-12pt}
\end{figure}

\subsection{The OBOE Observational Model}

\figref{fig:oboe} shows the main modeling constructs of OBOE (see:
\url{http://ecoinformatics.org/oboe/oboe.1.0/oboe-core.owl}). An {\em
  observation} is made of an {\em entity} (e.g., biological organisms,
geographic locations, environmental features) and serves to group a
set of measurements together to form a single ``\emph{observation
  event}''. A \emph{measurement} assigns a value to a {\em
  characteristic} of the observed entity (e.g., the weight of a
plant), and can also include \emph{standards} (e.g., units as well as
standards for coded values) and collection \emph{protocols}. An
observation can occur within the surrounding \emph{context} of other
observations (e.g., as part of a temporal or spatial context), and
context may include a named relationship (e.g., ``partOf'',
``within'') that existed during the observation event. A key feature
of OBOE is that it allows properties (characteristics and
relationships) of entities to be asserted without being interpreted as
\emph{inherently} (i.e., {always}) true of the entity.  Depending on
the context in which the entity was observed or how the measurements
were performed, an entity's properties may take on different values.
OBOE allows RDF-style assertions about entities to be contextualized,
and thus different values can be assigned for the same entity under
distinct contexts, which is a crucial feature for modeling ecological
as well as many other types of scientific data
\cite{bowers08,mungall07:_repres_phenot_in_owl}. In addition, OBOE is
currently implemented as an OWL-DL ontology that can be easily used
with (or extended by) other ontologies for specifying domain-specific
types of entities, characteristics, measurement standards, protocols,
and relationships. For instance, the Semtools project has defined
specific OBOE extensions in collaboration with the Santa Barbara
Coastal Long-Term Ecological Research Project as well as through
ongoing collaborations with other projects, and general extensions
exist for OBOE that define a number of common entities, measurement
units, and corresponding physical characteristics.

@@

\subsection{Semantic Annotations}

A semantic annotation consists of two parts: (i) a ``configuration''
of the observation model containing the specific entities,
characteristics, observations, measurements, and so on (drawn from one
or more domain ontologies) that appropriately capture the semantics of
the data set; and (ii) a mapping between the attributes in the data
set to specific measurements defined in the model
configuration. \figref{fig:kelp-mass-model} shows a high-level example
of an annotation defined for a simple Kelp sampling data set. Here,
the data set consists of five attributes (bottom of
\figref{fig:kelp-mass-model}). Each attribute is mapped to a specific
measurement type (where only the characteristic of each measurement
type is shown), and measurement types are organized into observations
of specific Kelp entities (shown of type ``Macrocystis''), temporal
points (denoted by date-times), and spatial locations (given as site
names). Each measurement associated with a Kelp observation is assumed
to have occurred within the site and during the given time as
specified by the context relationships.

@@

Semantic annotations can be used to facilitate discovery and
integration of heterogeneous data sets. For instance, combining
semantic annotations with OBOE, it is possible to discover data sets
based on searches expressed over types of observations and
measurements of interest. As simple examples, users can pose queries
such as ``\emph{find all data sets containing observations of Kelp}''
and ``\emph{find all data sets containing Mass measurements of
  Kelp}''.  Both of these queries would return the example data set in
\figref{fig:kelp-mass-model} since the attribute {\tt WET} is linked
to a WetMass measurement for observations of Macrocystis (where
WetMass is defined as a special kind, or \emph{subclass} of Mass, and
Macrocystis is defined as a subclass of Kelp). Using semantic
annotations in this way can help to increase both query recall and
precision over standard keyword-based approaches
\cite{berkley09:_improv_data_discov_for_metad}. In particular, by
defining terms as subclasses of other terms (e.g., Macrocystis as a
subclass of Kelp), term expansion can be used to increase the number
(recall) of data sets returned (where subclasses of query terms are
also searched). The precision of the result can be improved since
queries may specify the desired connections between terms (e.g.,
measurements of Mass for Kelp observations) as opposed to returning
all data sets that simply mention the terms but without any explicit
connections (i.e., where Mass was measured, but not for Kelp samples).
% Queries can additionally include context constraints, e.g., Mass
% measurements of Kelp within a specific spatial or temporal context.
Annotations also help facilitate integration by allowing
tools to align data set attributes based on their declared measurement
and observation types.

@@

\subsection{The Semantic Mediation API}

The Semantic Mediation API includes basic ontology management
features, annotation manipulation capabilities, and simple concept
navigation and visualization components. The API is intended to be a
centralized toolkit for use in multiple application contexts (on
either client or server deployments).  The Semantic Mediation API uses
both the OWL API \cite{owlapi} for ontology management services
including ontology parsing, serialization, and simple class and
property exploration as well as the Pellet description-logic reasoner
\cite{pellet} for classification and exposing inferred axioms in
source ontologies. The inference services exposed through the Semantic
Mediation API are used in both discovery and integration features
described below.
% is particularly powerful when utilizing OBOE Measurement 'templates'
% that express exactly what (Entity and Characteristic) can be
% observed and how (Standard and Protocol).
In our current Morpho and Metacat extensions, semantic annotations are
managed and stored automatically in an underlying, local relational
database. While it is also possible to use in-memory approaches for
storing and querying annotations, we found the overhead to be
prohibitive when large numbers of data sets are managed.

@@

\subsection{The Morpho Editor Plugin}

The semantic-annotation editor plugin for Morpho provides a front-end
to the Semantic Mediation API and allows data owners and curators to
define annotations for existing EML data descriptions. The editor
provides a simple ``fill-in-the-blank'' style form-based interface
with a searchable hierarchical concept selection widget (see
\figref{fig:morpho-annotation}). The plugin seamlessly integrates with
a standard Morpho installation and provides semantic query
capabilities for locating data packages, marking up data sets within a
package using semantic annotations, and saving annotations locally or
to a shared repository where they can be discovered and explored by
other users. The annotation editor in Morpho allows a user to view the
data set being annotated as they fill in (by selecting an appropriate
ontology term) the characteristic, measurement standard, protocol, and
associated entity for each data set attribute. Users can also specify
whether an observation spans multiple columns, and can provide context
relationships between attributes (i.e., observations). The editor
provides a number of additional features including the ability to view
the entire annotation (similar to \figref{fig:kelp-mass-model}) and to
specify additional mapping constraints for observations and
measurements.

\subsection{Metacat Query Extensions}

The semantic plugin for Metacat augments Metacat's existing metadata
storage and search by allowing annotations to be saved and queried
alongside the metadata and data that they annotate. In addition to
traditional keyword and spatial search criteria, the Metacat plugin
allows semantic criteria to be included where they may either increase
query recall using term-expansion (i.e., traversing the class
subsumption hierarchy) or refine the result set by limiting matches to
data sets that contain the specified observational model (e.g.,
combinations of OBOE-compatible entity, characteristic, measurement
standard, or protocol concepts). The observational model can be
leveraged further by materializing the annotation and data artifact
(via the Data Manager Library
\cite{leinfelder10:_metad_driven_approac_to_loadin}) into a fully
instantiated OBOE model and inspecting (and querying over) the
observational values themselves.

### 2011-07-30 docs/pubs/EIM-2011/intro.tex
snapshot copy -- will finish up rest later

A major challenge in environmental information management concerns
providing effective approaches for the discovery and integration of
heterogeneous data sets. For instance, locating and combining relevant
observational data are often critical and time-consuming steps for
researchers studying phenomena at broad spatial, temporal, and
biological scales
\cite{worm06:_impac_of_biodiv_loss_ocean_ecosy_servic,pennings05:_do,green05:_compl_in_ecolog_and_conser,sorokina09:_detec_inter_variab_inter_obser_ornit_data,jackson01:_histor_overf_and_recen_collap}. The
underlying data sets used within such studies frequently differ in
subtle and complex ways, due in part to the protocols used for data
collection, the types of observations made, and the experimental and
other contextual information associated with the data set. These
differences in turn can lead to structural and semantic heterogeneity
among data sets that make them hard to discover using current data
management approaches and require considerable manual effort by
researchers needing to combine data sets.

A number of recent efforts within the earth and environmental
informatics communities are adopting the notion of an \emph{observation} as a
key modeling concept for enabling improved discovery and integration
of scientific data
\cite{om,fox09:_ontol,tarboton07:_cuahs_commun_obser_data_model,cushing07:_compon_based_end_user_datab,balhoff10:_phenex,bowers08}. These
approaches provide higher-level observational data models for
describing and representing observations and measurements found in
underlying data sets by defining common ``core'' concepts such as the
entities or features being observed, measurement units and protocols,
and context relationships between observations \cite{om,bowers08}.  A
major goal of these approaches is to enable interoperability and
uniform access to data by abstracting away the underlying
representation details that often impede integration across scientific data sets.

>>>>

A major challenge in environmental information management concerns
providing effective approaches for the discovery and integration of
heterogeneous data sets. For instance, locating and combining relevant
observational data are often critical and time-consuming steps for
researchers studying phenomena at broad spatial, temporal, and
biological scales
\cite{worm06:_impac_of_biodiv_loss_ocean_ecosy_servic,pennings05:_do}. The
underlying data sets used within such studies frequently differ in
subtle and complex ways, due in part to the protocols used for data
collection, the types of observations made, and the experimental and
other contextual information associated with the data set. These
differences in turn can lead to structural and semantic heterogeneity
among data sets that make them hard to discover using current data
management approaches and require considerable manual effort by
researchers needing to combine data sets.

A number of recent efforts within the earth and environmental
informatics communities are adopting the notion of an
\emph{observation} as a key modeling concept for enabling improved
discovery and integration of scientific data
\cite{om,fox09:_ontol,tarboton07:_cuahs_commun_obser_data_model,mungall07:_repres_phenot_in_owl,bowers08}. These
approaches provide higher-level observational data models for
describing and representing observations and measurements found in
underlying data sets by defining common ``core'' concepts such as the
entities or features being observed, measurement units and protocols,
and context relationships between observations \cite{om,bowers08}.  A
major goal of these approaches is to enable interoperability and
uniform access to data by abstracting away the underlying
representation details that often impede integration across scientific
data sets.

### 2011-07-30 docs/pubs/EIM-2011/related.tex
further refinements: -use WET mass attribute in text since it appears in the figures/screenshots -finesse query example section, break long sentences into two -resolve DML citation link (was broken)
snapshot copy -- will finish up rest later

The need for more semantic mechanisms to describe observational data
has led to many proposals for observational data models (e.g.,
\cite{om,
  tarboton07:_cuahs_commun_obser_data_model,netcdf,cushing07:_compon_based_end_user_datab})
and ontologies (e.g.,
\cite{fox09:_ontol,sweet,sheth08:_seman_sensor_web,mungall110:_integ}).
The work presented here is complementary to these efforts by providing
a concrete set of software components that have been integrated with
popular metadata tools (namely, Metacat
\cite{metacat02:_manag_heter_ecolog_data_using_morph} and Morpho
\cite{berkley01:_metac}) to provide a more uniform, semantic view of
heterogeneous observational data. By extending Morpho and Metacat to
support semantic annotations, these tools can provide additional help
to researchers interested in performing synthetic studies by providing
semantically-enhanced discovery and integration services, which are
largely lacking in many existing environmental information management
frameworks \cite{jones_new_2006}.

Our work on using semantic annotations for data integration is closely
aligned to traditional information integration approaches (e.g.,
\cite{HalevyDataIntegration06,kolaitis05}), where a global mediated
schema is used to (physically or logically) merge the structures of
heterogeneous data sources using mapping constraints among the source
and target schemas. As such, the observational model we employ in our
framework can be viewed as a (general-purpose) mediation schema for
observational data sets.  This schema can be augmented with logic
rules (as target constraints) where semantic annotations are used as
mapping constraints. However, instead of users specifying logic
constraints directly, we provide a high-level annotation language and
user-interface components (through Morpho) that can simplify the
specification of mappings and more naturally aligns with the
observation model.

Finally, the use of annotations are playing a more prominent role in
database systems (e.g.,
\cite{GeertsKM06,Reeve05SemanticAnnotation,BhagwatCTV04}), e.g., the
MONDRIAN system \cite{GeertsKM06} employs an annotation model and a
set of query operators to manipulate both data and annotations.
However, users must be familiar with the underlying data structures
(schemas) to take advantage of these operators, which is generally not
feasible for observational data in which data sets exhibit a high
degree of structural and semantic heterogeneity. Our annotation
approach used to extend EML is also similar in spirit to a number of
other high-level mapping languages used for data exchange (e.g.,
\cite{fagin09:_clio,an06:_build_seman_mappin_datab_ontol}). Our
approach differs by being specifically tailored to the OBOE
observational model, which in turn simplifies the annotation language,
making it in general easier for users to specify annotations for
observational data. Our approach also provides well-defined and
unambiguous mappings from data sets to the observation and measurement
model, which is critical for providing automated, high-quality data
integration services over heterogeneous observational data.

>>>>

The need for more semantic mechanisms to describe observational data
has led to many proposals for observational data models (e.g.,
\cite{om, tarboton07:_cuahs_commun_obser_data_model,netcdf}) and
ontologies (e.g.,
\cite{fox09:_ontol,sweet,mungall07:_repres_phenot_in_owl}).  The work
presented here is complementary to these efforts by providing a
concrete set of software components that have been integrated with
popular metadata tools (namely, Metacat
\cite{metacat02:_manag_heter_ecolog_data_using_morph} and Morpho
\cite{berkley01:_metac}) to provide a more uniform, semantic view of
heterogeneous observational data. By extending Morpho and Metacat to
support semantic annotations, these tools can provide additional help
to researchers interested in performing synthetic studies by providing
semantically-enhanced discovery and integration services, which are
largely lacking in many existing environmental information management
frameworks \cite{jones_new_2006}.

Our work on using semantic annotations for data integration is closely
aligned to traditional information integration approaches (e.g.,
\cite{kolaitis05}), where a global mediated
schema is used to (physically or logically) merge the structures of
heterogeneous data sources using mapping constraints among the source
and target schemas. As such, the observational model we employ in our
framework can be viewed as a (general-purpose) mediation schema for
observational data sets.  This schema can be augmented with logic
rules (as target constraints) where semantic annotations are used as
mapping constraints. However, instead of users specifying logic
constraints directly, we provide a high-level annotation language and
user-interface components (through Morpho) that can simplify the
specification of mappings and more naturally aligns with the
observation model.

Annotations are playing a more prominent role in database systems,
e.g., the MONDRIAN system \cite{GeertsKM06} employs an annotation
model and a set of query operators to manipulate both data and
annotations.  However, users must be familiar with the underlying data
structures (schemas) to take advantage of these operators, which is
generally not feasible for observational data in which data sets
exhibit a high degree of structural and semantic heterogeneity. Our
annotation approach used to extend EML is also similar in spirit to a
number of other high-level mapping languages used for data exchange
(e.g., \cite{fagin09:_clio,an06:_build_seman_mappin_datab_ontol}). Our
approach differs by being specifically tailored to the OBOE
observational model, which in turn simplifies the annotation language,
making it in general easier for users to specify annotations for
observational data. Our approach also provides well-defined and
unambiguous mappings from data sets to the observation and measurement
model, which is critical for providing automated, high-quality data
integration services over heterogeneous observational data.
